---
title: "Qn6 - 9"
author: "Tilak Poudel"
date: "2025-04-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Q.6

## a. Define integers from 1 to 15 using three diffrent coding approaches in R
### 1. Using the colon operator
``` {r}
R <- 1:15
R
```

### 2. Using the seq() function

```{r}
R <- seq(1, 15)
R
```

### 3. Using the c() function
```{r}
R <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15)
R
```

## b Add these five numbers 1.1, 2.2, 3.3, 4.4, and 5.5 and save it as column vector

```{r}
numbers <- c(1.1, 2.2, 3.3, 4.4, 5.5)
N <- sum(numbers)
N
```


## c Add subtract , multiply divide vector R from vector N
```{r}
R_add <- R + N
R_add

R_sub <- R - N
R_sub

R_mul <- R * N
R_mul

R_div <- R / N
R_div

```

### d Define list
```{r}
L <- list("This", "is", "my", "first", "programming", "in", "R")
L
```

# Q.7  Import airquality dataset as dq object in R and do the following
## a. check the structure of the dataset and explain class of each variable
```{r}
aq <- airquality
aq
str(aq)
```
There are 153 observations and 6 variables in the dataset. The variables are:
- Ozone: numeric
- Solar.R: numeric
- Wind: numeric
- Temp: numeric
- Month: integer
- Day: integer

## b. Explain how to handle missing values of two variables aq
```{r}
# check missing values
is.na(aq)

## Ozone and Solar.R has missing value, handle missing value filling with mean
aq$Ozone[is.na(aq$Ozone)] <- mean(aq$Ozone, na.rm = TRUE)
aq$Solar.R[is.na(aq$Solar.R)] <- mean(aq$Solar.R, na.rm = TRUE)
aq
```

## c. Get the summary of all the variables and interpet the,
```{r}
    summary(aq)
```
The following is the summary (minimum, 1st quarter, median, mean 3rd quartle, max) for each variable with 
Ozone           Solar.R           Wind             Temp      
 Min.   :  1.00   Min.   :  7.0   Min.   : 1.700   Min.   :56.00  
 1st Qu.: 21.00   1st Qu.:120.0   1st Qu.: 7.400   1st Qu.:72.00  
 Median : 42.13   Median :194.0   Median : 9.700   Median :79.00  
 Mean   : 42.13   Mean   :185.9   Mean   : 9.958   Mean   :77.88  
 3rd Qu.: 46.00   3rd Qu.:256.0   3rd Qu.:11.500   3rd Qu.:85.00  
 Max.   :168.00   Max.   :334.0   Max.   :20.700   Max.   :97.00  

Month            Day      
 Min.   :5.000   Min.   : 1.0  
 1st Qu.:6.000   1st Qu.: 8.0  
 Median :7.000   Median :16.0  
 Mean   :6.993   Mean   :15.8  
 3rd Qu.:8.000   3rd Qu.:23.0  
 Max.   :9.000   Max.   :31.0 

## d. Get the summary statistics of temp variable by month categories and interpret

```{r}
summary(aq$Temp, aq$Month)
```
The following is the interpretation of temp by month
 Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  56.00   72.00   79.00   77.88   85.00   97.00 

The minimum temp is 56 , max as 97 , 79 is the median and 77.88 as mean.


# QN 8. 
## a. Create a bar plot of Month variable and interpret 
```{r}
barplot(table(aq$Month), 
  main = "Bar plot of Month", 
  xlab = "Month", 
  ylab = "Frequency", 
  col = "blue")
```
### Interpretation
  - The bar plot shows the frequency of each month in the dataset.
  - The month with the highest frequency is 7, followed by 5 and 8. 
  - The month with the lowest frequency is 6 and 9.

## b. Create a histogram of Temp variable and interpret
```{r}
hist(aq$Temp, 
  main = "Histogram of Temp", 
  xlab = "Temperature", 
  ylab = "Frequency", 
  col = "green"
)
```
Interpretation,
- - The histogram shows the distribution of temperature values in the dataset.
- The temperature values are concentrated around 75 to 85 degrees, with a peak around 75 degrees.
- The distribution is slightly right-skewed, indicating that there are some higher temperature values.
- The histogram has a normal distribution shape, indicating that the temperature values are normally distributed.

## c. Create line plot of Temp and Day and interpet
```{r}
plot(aq$Day, aq$Temp, 
  type = "l", 
  main = "Line plot of Temp and Day", 
  xlab = "Day", 
  ylab = "Temperature", 
  col = "red"
)
```
### Interpretation
- The line plot shows the relationship between temperature and day.
- The temperature values increase from day 1 to day 31, with some fluctuations in between.
- The line plot doesnot distinctlively indicate the correlation.

## d. Create a scatter plot of Ozone and Solar and interpret
```{r}
plot(aq$Ozone, aq$Solar.R, 
  main = "Scatter plot of Ozone and Solar", 
  xlab = "Ozone", 
  ylab = "Solar", 
  col = "purple"
)
```
### Interpretation
- The scatter plot shows the relationship between ozone and solar radiation.  
- There is a positive correlation between ozone and solar radiation, indicating that as ozone levels increase, solar radiation also increases.
- The scatter plot also shows some outliers, indicating that there are some extreme values in the dataset.

# QN 9.Do the following with tidyverse package
### a. Define tibble having country, year, cases and population with 20 random values in each
```{r}
#library(tidyverse)
library(tibble)
set.seed(39)
country <- sample(c("USA", "India", "China", "Brazil", "UK"), 20, replace = TRUE)
year <- sample(2000:2020, 20, replace = TRUE)
cases <- sample(1000:10000, 20, replace = TRUE)
population <- sample(1000000:10000000, 20, replace = TRUE)
tibble_data <- tibble(country, year, cases, population)
tibble_data
```

### b. Transform the cases as log of cases(LnCase) and population as log of population(LnPop)
```{r}
library(dplyr)
tibble_data <- tibble_data %>%
  filter(cases > 0, population > 0) %>%  # Remove non-positive values
  mutate(LnCase = log(cases), 
         LnPop = log(population))
tibble_data
```

### c. Create a scatter plot of 
#### c1. cases and population

```{r}
plot(tibble_data$cases, tibble_data$population, 
     main = "Scatter plot of Cases and Population", 
     xlab = "Cases", 
     ylab = "Population", 
     col = "blue")
```

#### c2. LnCase and population
```{r}
plot(tibble_data$LnCase, tibble_data$population, 
     main = "Scatter plot of LnCase and Population", 
     xlab = "LnCase", 
     ylab = "Population", 
     col = "red")
```
# Error: attempt to use zero-length variable name

#### c3. cases and LnPop
```{r}
  plot(tibble_data$cases, tibble_data$LnPop, 
      main = "Scatter plot of Case and LNPopulation", 
      xlab = "Cases", 
      ylab = "Population", 
      col = "blue")
```

#### c4. LnCase and LnPop
```{r}
    plot(tibble_data$LnCase, tibble_data$LnPop, 
      main = "Scatter plot of Case and LNPopulation", 
      xlab = "Cases", 
      ylab = "Population", 
      col = "blue")
```
  
### d. Show four graphs in a single graph window
```{r}
#Creat a window for graphs in 2 rows and 2 columns
par(mfrow = c(2, 2))
plot(tibble_data$cases, tibble_data$population, 
     main = "Scatter plot of Cases and Population", 
     xlab = "Cases", 
     ylab = "Population", 
     col = "blue")
plot(tibble_data$LnCase, tibble_data$population, 
     main = "Scatter plot of LnCase and Population", 
     xlab = "LnCase", 
     ylab = "Population", 
     col = "red")
  plot(tibble_data$cases, tibble_data$LnPop, 
     main = "Scatter plot of Case and LNPopulation", 
     xlab = "Cases", 
     ylab = "Population", 
     col = "green")
  plot(tibble_data$LnCase, tibble_data$LnPop, 
     main = "Scatter plot of Case and LNPopulation", 
     xlab = "Cases", 
     ylab = "Population", 
     col = "purple")
# Graph is default mode
# par(mfrow = c(1, 1))
```
