---
title: "self-practice"
output:
  pdf_document: default
  html_document: default
---

```{r}
gd <- as.data.frame(VADeaths)
gd

# changing axis lenght and font-size
barplot(
    gd$`Rural Male`,
    horiz = TRUE,
    names.arg = c("50-54", "55-59", "60-64", "65-69", "70-74"),
    main = "Death Rates in Virginia",
    xlab = "Age Groups",
    ylab = "Death Rates",
    col = "lightblue",
    border = "red",
    beside = TRUE,
    xlim = c(0, 70),
    cex.axis = 0.8,
    cex.names = 0.8
)
```

## Sub divided bar plot
```{r}
gdm <- as.matrix(gd)
gdm
barplot(
    gdm,
    main = "Death Rates in Virginia",
    xlab = "Age Groups",
    ylab = "Death Rates",
    col = c("lightblue", "lightgreen", "lightpink"),
    border = "red",
    legend = rownames(gd),
)
```
## Multiple / Grouped bar diagram

```{r}
barplot(
    gdm,
    col = c("lightblue", "mistyrose", "lightcyan", "lavender", "cornsilk"),
    legend  = rownames(gdm),
    beside = T
)
#  NOTE: Adding beside = TRUE will produce the multiple bar chart
```
## Multiple / Group Bar diagram with change in legend values
```{r barplot-with-legend, echo=TRUE}
gdm <- as.matrix(VADeaths)

my_colors <- c("lightblue", "mistyrose", "lightcyan", "lavender", "cornsilk")
barplot(
    gdm,
    col = my_colors,
    beside = TRUE,
    main = "Death Rates in Virginia",
    xlab = "Age Groups",
    ylab = "Death Rates",
    cex.axis = 0.8,
    cex.names = 0.8
)
legend(
    "topleft", 
    legend = rownames(gdm), 
    fill = my_colors, 
    cex = 0.8, 
    lty = 0
)
```

## Pie chart
```{r}
gd <- as.data.frame(VADeaths)
pie(
    gd$`Rural Male`,
    main = "Death Rates in Virginia",
    labels = rownames(gd),
    radius = 1,
    col = c("lightblue", "lightgreen", "lightpink", "lightyellow", "lightgray"),
)
```

## Pie chart with percentage
```{r}
gd$piepercent <- round(gd$`Rural Male` / sum(gd$`Rural Male`) * 100, 1)
pie(
    gd$`Rural Male`,
    main = "Death Rates in Virginia",
    labels = paste(rownames(gd), "\n", gd$piepercent, "%"),
    radius = 1,
    col = rainbow(length(gd$`Rural Male`)),
)
# How to place a legend on the topright corner?
# Pie chart with percentage and legend
legend(
    "topright", 
    legend = rownames(gd), 
    fill = rainbow(length(gd$`Rural Male`)), 
    cex = 0.8, 
    lty = 0
)
```

## Historgram
```{r}
gdcar = as.data.frame(cars)
str(gdcar)
plot(gdcar$speed)
```

## Density plot with polygon fill: Speed variable
```{r}
dens <- density(cars$speed)
plot(
    dens,
    main = "Density plot of Speed(mpg)",
    xlab = "Speed",
    ylab = "Density",
    col = "steelblue",
    frame = FALSE
)
polygon(dens, col = "lightblue", border = "black")
```

## Q-Q plot
### A Q-Q plot (quantile-quantile plot) is a graphical tool to help assess if a dataset follows a given distribution.
# Always use this plot to access normality
### Q-Q plot of speed variable
```{r}
qqnorm(gdcar$speed)
qqline(gdcar$speed, col = "red")
```

# Scatter plot
```{r}
# Scatterplot
plot(
    gdcar$speed,
    gdcar$dist,
    main = "Scatterplot of Speed vs Distance",
    xlab = "Speed",
    ylab = "Distance",
    col = "steelblue",
    pch = 19
)
```
# Boxplot
## A boxplot is a standardized way of displaying the distribution of data based on a five-number summary ("minimum", first quartile (Q1), median, third quartile (Q3), and "maximum").
```{r}
boxplot(
    mpg~cyl, 
    data=mtcars,
    main = "Boxplot of mpg by cyl",
    xlab = "Number of cylinders",
    ylab = "Miles per gallon",
    col = "lightblue",
    border = "red",
    horizontal = FALSE,
    notch = FALSE
)

```