---
title: "39_Project1"
author: "Tilak Poudel"
date: "2025-03-09"
output: html_document
---

## Session 1 & 2

Example_1
Column vector
```{r}
data<-c(1,2,3,4,5,6,7,8,9)
data
text<-c("a","b","c","d","e","f","g","h","i")
text
data2<-cbind(data,text)
data2
```

Example_2

Matrix contains in 2 dimensions in table like structure
```{r}
M<-matrix(
  c(1:9), nrow = 3, ncol = 3, byrow = TRUE
)
print(M)
```

Example_3
```{r}
V<-c(1:12)
MDA<-array(V,dim = c(2,3,2))
print(MDA)
```

## Data frames in R
### Creating data frame in R.

```{r}
df <- data.frame(x=c(1,2,3), y=c(10,20,30), z=c(30,45,50))
df

class(df)
```

Create a data frame

```{r}
emp.data <- data.frame(
  emp_id = c(1:5),
  emp_name =c("Rick", "Dan", "Michelle", "Ryan", "Gary"),
  salary = c(623.3, 515.2, 611.0, 729.0, 845.25),
  start_date = as.Date(c("2012-01-01", "2013-09-23", "2014-11-15", "2014-05-11", "2015-03-27")),
  stringsAsFactors = FALSE
)

# print the data
print(emp.data)

# Get the structure of the data frame.
str(emp.data)

# Print the summary
print(summary(emp.data))

# Extract part of data from data frame
# Extract emp_name and salary
result <- data.frame(emp.data$emp_name, emp.data$salary)
print(result)

# Extract first two rows
result1 <- emp.data[1:2, ]
print(result1)

# Extract 3rd and 5th row with 2nd and 4th column
result <- emp.data[c(3,5), c(2,4)]
print(result)

# Add new column in existing data frame
# Add "dept" column
emp.data$dept <- c("IT", "HR", "Finance", "Operations", "IT")
v <- emp.data

print(v)

# Expand data frame in R (Adding cases)
# Create a new data frame
emp.newdata <- data.frame(
  emp_id = c(6:8),
  emp_name =c("Ram", "Shyam", "Mohan"),
  salary = c(923.3, 715.2, 681.0),
  start_date = as.Date(c("2014-01-01", "2013-07-20", "2014-11-25")),
  dept = c("HR", "IT", "Finance"),
  stringsAsFactors = FALSE
)

print(emp.newdata)

# Expand(Combine) data frame using rbind 
# Bind the tow data frames
emp.finaldata <- rbind(emp.data, emp.newdata)
print(emp.finaldata)
```

## Import data in R
```{r}

# load necessary libraries
library(magrittr)

# Read csv file
covid_nep_data = read.csv('../data/covnep_252days.csv')

print(str(covid_nep_data))

print(head(covid_nep_data))

print(summary(covid_nep_data))
```
Importing COVID data in Text Files
```{r}
y<-read.csv("../data/covnep_252days.csv")
head(y)
```
Importing data into Excel files
```{r}
library(readxl)
covid_data_xls <- readxl::read_excel("../data/covnep_252days.xlsx")
tail(covid_data_xls)
```

## Session 3
### Basics of R
#### PEMDAS Rule in R
```{r}
s<-(4+6)^2*5/10+9-1
print(s)
```

#### Data types in R

1) Integer data type
```{r}
x<-c(1,2,3,4,5,6,7,8,9)
class(x)
```

```{r}
y<-c(1L,2L,3L,4L,5L,6L,7L,8L,9L)
print(y)
class(y)
```

R promotes integer to numeric when needed

(Multiplying integer by numeric)
```{r}
z<-4L*2.8
print(z)
```

Dividing integer by integer will give numeric
```{r}
b<-5L/2L
print(b)
```

Data_Types_In_R
1) Character
```{r}
x<-"data"
print(x)
class(x)
```

2)Factor
```{r}
y<-factor("data")
print(y)
class(y)
```
Factor and attributes in R
```{r}
gender<-factor(c("male","female","female","male"))
typeof(gender)
attributes(gender)
```

```{r}
unclass(gender)
```
Data_Types

Date: To store Date
```{r}
date1<-as.Date("2025-03-09")
date1
```

```{r}
class(date1)
```
POSIXct:To store date and time

```{r}
date2<-as.POSIXct("2025-03-09 12:57")
date2
```

```{r}
class(date2)

```
Logical_Data_Types

```{r}
x<- 2 == 3
x
```

```{r}
2!=3
```

```{r}
2<3
```

```{r}
2<=3
```

```{r}
2>3
```

```{r}
2>=3
```

```{r}
"data" == "stats"
```

```{r}
"data" < "stats"
```

```{r}
"hari" == "hare"
```
#### Vector operation in R(multiplication by a scalar)
```{r}
x1<-c(1,2,3,4,5,6,7,8,9)
y<-x1*3
y
```
Addition with Scalar
```{r}
z<-x1+2
z
```
Subtraction With Scalar
```{r}
d<-x1-2
d
```
Square root 
```{r}
e<-sqrt(x1)
e
```
Extending vector operation in R
(Two vectors of equal length)
```{r}
x<-1:10
y<- -5:4
length(x)
```

```{r}
length(y)
```
Addition
```{r}
x+y
```
Subtraction
```{r}
x-y
```
Multiplication
```{r}
x*y
```
Division

```{r}
x/y
```
Power
```{r}
x^y
```
Addition of both length
```{r}
length(x+y)
```
Two vector of unequal length
(Addition)
```{r}
x<-1:10
z<-c(1,2)
x+z
```
Accessing individual elements in R

```{r}
x[1]
```

```{r}
x[c(1,4)]
```
Creation and giving name to a Vector
```{r}
w1<-1:3
names(w1)<-c("a", "b", "c")
w1
```

#### Missing data in R (NA Type)
```{r}
zchar<-c("Hockey", NA, "Cricket")
is.na(zchar)
```

```{r}
nchar(zchar)
```
Calculating mean of missing Data
```{r}
z<-c(1,2,NA,8,3,NA,3)
mean(z)
```
Removing NA and  calculating mean
```{r}
mean(z, na.rm = TRUE)
```
Calculating variance excluding NA 
```{r}
var(z, na.rm = TRUE)
```
NULL type missing in R
```{r}
z<-c(1, NULL, 3)
z
```

```{r}
is.null(z)
```

```{r}
f<-NULL
is.null(f)
```

#### Pipes in R
```{r}
x<-1:10
mean(x)
```
With pipe operator(%>%)
```{r}
library(magrittr)
x%>%mean
```

```{r}
z<-c(1,2,NA,8,3,NA,3)
sum(is.na(z))
```
Pipes without nesting 
```{r}
z %>% is.na %>% sum
```
Additional argument

```{r}
z %>% mean(na.rm = TRUE)
```

### Advanced Data Structure In R
(Data Frame)

```{r}
x<-10:1
y<- -4:5
q<-c("Hockey","Football","Baseball","Kabaddi","Rugby","Pingpong","Basketball","Tennis","Cricket","Volleyball")
sportsDF<-data.frame(x,y,q)
sportsDF

```

Naming data frame
```{r}
sportsDF<-data.frame(First =x, Second = y, Sport = q)
sportsDF
```
Accessing column
```{r}
names(sportsDF)[3]
```

```{r}
rownames(sportsDF)
```

```{r}
rownames(sportsDF)<-c("One","Two","Three","Four","Five","Six","Seven","Eight","Nine","Ten")
sportsDF
```
Printing first few Rows
```{r}
head(sportsDF)
```

Printing last few rows
```{r}
tail(sportsDF)
```
Printing first seven rows
```{r}
head(sportsDF, n =7)
```
Accessing value from third row and second column
```{r}
sportsDF[3,2]
```
Accessing from Third column

```{r}
sportsDF[,3]
```

```{r}
sportsDF[,c("First","Sport")]
```

```{r}
sportsDF[,"Sport",drop = FALSE]
```
#### Lists in R
```{r}
list1<-list(1,2,3)
list1
```
Single element list
```{r}
list2<-list(c(1,2,3))
list2
```

```{r}
list3<-list(c(1,2,3),3:7)
list3
```
List with data frame and vector
```{r}
list4<-list(sportsDF,1:10)
list4
```

```{r}
list5<-list(sportsDF,1:10,list3)
list5
```

```{r}
names(list5)
```

```{r}
names(list5)<-c("dataframe","vector","list")
list5
```
Accessing elements of list
```{r}
list5[[1]]
```

```{r}
list5[["dataframe"]]
```
