---
title: '7'
author: "Utsab Bhattarai"
date: "2024-06-27"
output:
  pdf_document: default
  html_document: default
---

# Do the followings in R studio using "airquality" dataset with R script:
```{r}
# Load necessary packages
library(ggplot2)
library(dplyr)

# Load the airquality dataset
data("airquality")
```

## a. Get the boxplot of Temp variable using ggplot2 package and interpret it carefully.
```{r}
ggplot(airquality,
       aes(y = Temp)) + 
  geom_boxplot(fill = "skyblue") + 
  labs(title = "Boxplot of Temperature",
       y = "Temperature (°F)") +
  theme_minimal()

# Interpretation:
# The boxplot shows the distribution of the temperature variable. 
# The median temperature is represented by the thick line inside the box
# which is around 79(°F).
# The interquartile range (IQR) is shown by the height of the box,
# which represents the middle 50% of the data.
# The "whiskers" extend to 1.5 times the IQR from the quartiles.
# Any points outside the whiskers are considered outliers.
```

## b. Create class intervals of Temp variable using dplyr package and show it as frequency distribution.
```{r}
airquality <- airquality %>% 
  mutate(TempClass = cut(Temp,
                         breaks = seq(min(Temp),
                                      max(Temp),
                                      by = 5)))

temp_freq <- airquality %>%
  group_by(TempClass) %>%
  summarise(Frequency = n())

print(temp_freq)
```

## c. Get pie chart of Temp variable class intervals using ggplot2 package and interpret it carefully.
```{r}
ggplot(temp_freq,
       aes(x = "",
           y = Frequency,
           fill = TempClass)) +
  geom_bar(stat = "identity",
           width = 1) +
  coord_polar("y") +
  labs(title = "Pie Chart of Temperature Class Intervals",
       y = "Frequency",
       fill = "Temperature Class") +
  theme_minimal()

# Interpretation:
# The pie chart displays the distribution of temperature class intervals.
# Each slice represents a temperature class interval and its relative frequency.
# The size of each slice corresponds to the frequency of observations within that interval.
```

## d. Get scatterplot of corrected Temp and Wind variables using ggplot2 package and interpret it carefully.
```{r}
ggplot(airquality,
       aes(x = Wind,
           y = Temp)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm",
              se = FALSE,
              color = "red") +
  labs(title = "Scatterplot of Temperature vs Wind",
       x = "Wind (mph)",
       y = "Temperature (°F)") +
  theme_minimal()

# Interpretation:
# The scatterplot shows the relationship between temperature and wind speed.
# Each point represents a day's observation.
# The red line is a linear regression line that shows the trend between wind speed and temperature.
# There seems to be a negative relationship,
# indicating that higher wind speeds are associated with lower temperatures.
```